Vertical and oblique transmission with fluctuating transmission

### 2018-05-13 ms/ms.tex
rename
analytic results for migration model

\documentclass[12pt]{extarticle} %twocolumn
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

@@

Following migration, the frequencies of $A$ in the two demes are
\begin{equation} \label{eq:migration_model}
\begin{aligned}
x_1'' = (1-m_1) x_1' + m_2 x_2', \\
x_2'' = m_1 x_1' + (1-m_2) x_2',
\end{aligned}
\end{equation}
where $m_j$ is the migration rate out of deme $j$.

\autoref{fig:migration} shows the stable frequencies of phenotype $A$ (i.e. the frequencies that solve $x_1=x_1''$ and $x_2=x_2''$) and the stable population mean fitnesses in the two demes with symmetric selection such that $A$ is favored in deme 1 and $B$ is favored in deme 2 with similar intensities (see figure legend).
Notably, without oblique transmission (left column), migration decreases the differences between the demes and reduces the population mean fitnesses.
With some oblique transmission, but equal in both demes, results are similar (not shown).
However, if oblique transmission is stronger in deme 2 compared to deme 1 (middle and right columns), the stable frequency of $A$ increases in both demes.
Therefore, the mean fitness in deme 1 decreases to a lesser extend than in deme 2, and even increases when the migration rate is high enough.
Furthermore, the polymorphism that is protected when transmission rates are equal disappears when transmission rates differ and migration is strong enough, and phenotype $A$ fixes (see right side of right column).

\begin{figure*}[ht]
\centering
\includegraphics[width=0.75\linewidth]{../figures/migration.pdf}
\caption{
\textbf{Oblique transmission and migration.} 
The figure shows $x^*_i$ the stable frequency of $A$ (top row) and $\bar{w}^*_i$ population mean fitness (bottom row) in the two demes.
Selection is symmetric between the two demes (the fitness of phenotype $A$ is $w_1=2$ in deme 1 and $w_2=0.5$ in deme 2; the fitness of phenotype 2 is $1$ in both demes).
The vertical transmission rate is $\rho_1=1$ in deme 1, and $\rho_2=1$, $0.8$, and $0.25$, in the left, middle, and right columns, respectively.
Migration rate $m$ is on the x-axis.
}
\label{fig:migration}
\end{figure*}

@@

\bibliographystyle{agsm}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{fluctuating_oblique}

>>>>

\documentclass[11pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

@@

Following migration, the frequencies of $A$ in the two demes are
\begin{equation} \label{eq:migration_model}
\begin{aligned}
x_1'' &= (1-m_1) x_1' + m_2 x_2', \\
x_2'' &= m_1 x_1' + (1-m_2) x_2',
\end{aligned}
\end{equation}
where $0 \le m_j \le 1/2$ is the migration rate out of deme $j$.

For the simplified case of unconditional migration $m_1=m_2=m$ \citep{McPeek1992} and symmetric selection $w_1=1/w_2=w>1$, the equilibrium $(x_1^*, x_2^*)$ solves
\begin{equation}\begin{aligned} \label{eq:migration_model_unconditional_symmetric}
x_1 &= (1-m)x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + m x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big), \\
x_2 &= m x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + (1-m) x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big).
\end{aligned}
\end{equation}

With only oblique transmission ($\rho_1=\rho_2=0$), the solutions to \eqref{eq:migration_model_unconditional_symmetric} are the neutral equilibria $(x^*,x^*)$ for any $0 \le x^* \le 1$.
With only vertical transmission ($\rho_1=\rho_2=1$), the solutions to \eqref{eq:migration_model_unconditional_symmetric} are $(0,0)$, $(1,1)$ and the polymorphic solution
\begin{equation}
\begin{aligned} \label{eq:migration_model_unconditional_symmetric_solution}
x_1^* &= \frac{w-1-m(w+1) + \sqrt{\Delta}}{2(w-1)}, \\
x_2^* &= 1-x_1^*,
\end{aligned}
\end{equation}
where $\Delta = m^2(w+1)^2+(1-2m)(w-1)^2$.
Examining the stability of these equilibria, the characteristic polynomial of the Jacobian is the same for both fixation equilibria $(0,0)$ and $(1,1)$ and is given by $a \lambda^2 + b \lambda + c$ with
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{(1-m)(w^2+1)}{w}, \quad
c &=  1-2m.
\end{aligned}
\end{equation}
Since $a>0$, $b<0$, and $c>0$, and because $a+b+c=-\frac{1-m}{w}(w-1)^2<0$, the leading eigenvalue is positive, both equilibria are unstable, and the protected polymorphism exists and is given by~\eqref{eq:migration_model_unconditional_symmetric_solution}.

With only vertical transmission in one deme ($\rho_1=1$) and a combination of both vertical and oblique transmission in the other deme ($\rho_2=\rho$), the stability of $(0,0)$ is determined by a characteristic polynomial with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{(1-m)(w+1-\rho+\rho/w)}{w}, \quad
c &=  (1-2m)(w(1-\rho)+\rho).
\end{aligned}
\end{equation}
Again, $a>0$ and $c>0$   and stability can be analyzed by the sign of $a+b+c$.
For $\rho=1$ we already have $a+b+c<0$.
For $\rho=0$ we have $a+b+c=-m(w-1)<0$.
Finally, $a+b+c$ is a linear function of $\rho$ and therefore $a+b+c<0$ for any valid $\rho$, the leading eigenvalue is positive, and $(0,0)$ is unstable.
Regarding the stability of $(1,1)$, which is determined by a characteristic polynomial  with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{1-m}{w}(1 + w(1-\rho) +w^2 \rho), \quad
c &= \frac{1-2m}{w} (w\rho+1-\rho).
\end{aligned}
\end{equation}
Once more, $a>0$, $b<0$, and $c>0$, and if $\rho=1$ then $a+b+c<0$, as before.
However, for $\rho=0$ we have $a+b+c = \frac{m}{w}(w-1)>0$, in which case the stability is determined by $2a+b=\frac{m(w+1)+w-1}{w}$, the derivative of the polynomial at $\lambda=1$, which is positive, and therefore both eigenvalues occur at $(0,1)$ and the fixation is stable.
We can also determine the rate $\hat \rho$ for which $a+b+c=0$, such that for
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_rho}
\rho < \hat \rho = \frac{m}{m+(1-m)(w-1)}
\end{equation}
the fixation $(1,1)$ is stable if and only if $2a+b>0$.
We already found that when $\rho=0$ we have $2a+b>0$.
When $\rho = \hat \rho$, we have $2a+b=\frac{m(w+1)+w-1}{w(1-m)(w-1)}>0$.
Since $2a+b$ is linear in $\rho$, we find that $2a+b>0$ for $\hat \rho \le \rho \le 0$ and therefore \eqref{eq:migration_model_unconditional_symmetric_condition_rho} guarantees the stable fixation of $A$.

A similar condition can be found for $m$: the value of $a+b+c$ at $m=0$ and $m=1$ is $-\frac{\rho(w-1)^2}{w}<0$ and $\frac{(1-\rho)(w-1)}{w}>0$, respectively, and $a+b+c$ is linear in $m$, so that we have $a+b+c>0$ when 
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_m}
m > \hat m = \frac{\rho (w-1)}{\rho (w-1) +1-\rho}.
\end{equation}
In addition, $2a+b$ is linear in $m$ and equal to 2 when $m=1$; it is equal to zero when $m=(w-1)(w\rho-1)/(w(w-1)\rho+w+1) < \hat m$, and therefore \eqref{eq:migration_model_unconditional_symmetric_condition_m} is sufficient for stability of fixation of $A$.

\autoref{fig:migration_rho} and \autoref{fig:migration_m} show the stable frequencies of phenotype $A$ (i.e. the frequencies that solve \eqref{eq:migration_model_unconditional_symmetric_solution}) and the stable population mean fitnesses in the two demes with symmetric selection such that $A$ is favored in deme 1 and $B$ is favored in deme 2 with similar intensities (see figure legend).
Notably, without oblique transmission (\autoref{fig:migration_m}, left column), migration decreases the differences between the demes and reduces the population mean fitnesses.
With some oblique transmission, but equal in both demes, results are similar (not shown).
However, if oblique transmission is stronger in deme 2 compared to deme 1 (\autoref{fig:migration_m}, middle and right columns), the stable frequency of $A$ increases in both demes.
Therefore, the mean fitness in deme 1 decreases to a lesser extent than in deme 2, and even increases when the migration rate is high enough.

The polymorphism $(x_1^*, x_2^*)$ is protected when transmission rates are equal, but not when transmission rates differ enough and migration is strong enough, and phenotype $A$ fixes. 
For example, with enough oblique transmission ($\rho_2<\hat \rho$) in deme 2, phenotype $A$ fixes, and the stronger migration is, the less oblique transmission is required to fix $A$, see shaded areas in \autoref{fig:migration_rho}.
Similarly, with enough migration ($m > \hat m$), phenotype $A$ fixes, and the more oblique transmission in deme 2, the less migration is needed to fix $A$; see shaded area in \autoref{fig:migration_m}.

\begin{figure*}[ht]
\centering
\includegraphics[width=0.75\linewidth]{../figures/migration_rho.pdf}
\caption{
\textbf{Oblique transmission and migration: effect of transmission.} 
The figure shows $x^*_i$ the stable frequency of $A$ (top row) and $\bar{w}^*_i$ population mean fitness (bottom row) in the two demes.
Selection is symmetric between the two demes (the fitness of phenotype $A$ is $w_1=2$ in deme 1 and $w_2=0.5$ in deme 2; the fitness of phenotype 2 is $1$ in both demes).
The vertical transmission rate is $\rho_1=1$ in deme 1, and $\rho_2$ (x-axis) in deme 2.
Migration rate $m$ is 0.05, 0.1, or 0.25 I the left, middle, and right columns, respectively.
The shaded area denotes stables fixation of phenotype $A$ according to~\eqref{eq:migration_model_unconditional_symmetric_condition_rho}.
}
\label{fig:migration_rho}
\end{figure*}

\begin{figure*}[ht]
\centering
\includegraphics[width=0.75\linewidth]{../figures/migration_m.pdf}
\caption{
\textbf{Oblique transmission and migration: effect of migration.} 
The figure shows $x^*_i$ the stable frequency of $A$ (top row) and $\bar{w}^*_i$ population mean fitness (bottom row) in the two demes.
Selection is symmetric between the two demes (the fitness of phenotype $A$ is $w_1=2$ in deme 1 and $w_2=0.5$ in deme 2; the fitness of phenotype 2 is $1$ in both demes).
The vertical transmission rate is $\rho_1=1$ in deme 1, and $\rho_2=1$, $0.4$, and $0.2$, in the left, middle, and right columns, respectively, in deme 2.
Migration rate $m$ is on the x-axis.
The shaded area denotes stables fixation of phenotype $A$ according to~\eqref{eq:migration_model_unconditional_symmetric_condition_m}.
}
\label{fig:migration_m}
\end{figure*}

@@

\bibliographystyle{agsm}
\bibliography{/Users/yoavram/Documents/library}
%\bibliography{ms}
